Vagueness, Noise, and Signaling

### 2015-09-29 paper/version_04/paper.tex
shortened section 3 to move technicalities into appendix; first stages

Formally, a sim-max game consists of a set of states $\States$, a set of messages $\Messgs$ typically with much
fewer messages than states, a probability distribution $\Pr \in \Delta(\States)$ such that
$\Pr(\state)$ gives the probability that state $\state$ occurs, a similarity metric\footnote{A
  metric is a function of distance between any two points in a given space. It should satisfy
  certain axioms that ensure behavior that one would intuitively expect from the words
  ``similarity'' and ``distance'' alone, but these details do not matter for the purposes of
  this paper.}
  % : similarity should always be non-negative, zero only for the distance
  % between an element and itself, symmetric, and satisfy the triangle inequality that
  % $\similarity(x,z) \le \similarity(x,y) + \similarity(y,z)$
on states $\similarity \mycolon \States \times \States \rightarrow \mathds{R}$ such that
$\similarity(\mystate{1}, \mystate{2})$ is the (perceptual) similarity between $\mystate{1}$
and $\mystate{2}$, and a utility function $\Utils \mycolon \States \times \States \rightarrow
\mathds{R}$ such that $\Utils(\mystate{1}, \mystate{2})$ is the payoff for sender and receiver
for a play with actual state $\mystate{1}$ and receiver interpretation $\mystate{2}$. We
identify the receiver's acts with the states of the world, so that the game is one of guessing
the actual state, so to speak. The utility function should be a monotonically decreasing
function of similarity.
A sender's behavior can be represented by a pure strategy $\Spure \in \Messgs^\States$, which deterministically defines which message would be used for each state, a mixed strategy $\Smixed \in \Delta(\Messgs^\States)$, which characterizes a probability distribution over pure strategies, or a behavioral strategy $\Sstrat \in \Delta(\Messgs)^\States$, which defines the \emph{probability} that a given message would be used for each state.
The analogous happens for the receiver's behavior.
A play in such a game with behavioral strategies is illustrated in Figure~\ref{fig:sim-max-illustration}.

@@

  \node (1) {};
  \node (2) [right=of 1,label=below:$\Pr(\mystate{1})$] {$\mystate{1} \in \States$};
  \node (3) [right=of 2,label=below:$\Sstrat(\messg \probbar \mystate{1})$] {$\messg \in \Messgs$};
  \node (4) [right=of 3,label=below:$\Rstrat(\mystate{2} \probbar \messg)$] {$\mystate{2} \in \States$};

@@

  \caption{A round of play in a sim-max game with behavioral strategies: Nature (N) chooses a state $\mystate{1}$ with probability $\Pr(\mystate{1})$; a sender (S) selects a message $\messg$ with probability $\Sstrat(\messg \probbar \mystate{1})$; a receiver (R) selects a state $\mystate{2}$ with probability $\Rstrat(\mystate{2} \probbar \messg)$. Payoff for both sender and receiver is given by $\Utils(\mystate{1}, \mystate{2})$.}
   %which is typically proportional to state similarity $\similarity(\mystate{1}, \mystate{2})$.}
  \label{fig:sim-max-illustration}
\end{figure}

@@

@@

>>>>

@@

  \node (1) {};
  \node (2) [right=of 1,label=below:$\Pr(\mystate{1})$] {$\mystate{1} \in \States$};
  \node (3) [right=of 2] {$\messg = \Spure(\mystate{1}) \in \Messgs$};
  \node (4) [right=of 3] {$\mystate{2} = \Rpure(\messg) \in \States$};

@@

  \caption{A round of play in a sim-max game: Nature (N) chooses a state $\mystate{1}$ with probability $\Pr(\mystate{1})$; a sender (S) selects a message $\messg = \Spure(\mystate{1})$ determined by his pure strategy; a receiver (R) selects a state $\mystate{2} = \Rpure(\messg)$ determined by her pure strategy. Payoff for both sender and receiver is given by $\Utils(\mystate{1}, \mystate{2})$.}
   %which is typically proportional to state similarity $\similarity(\mystate{1}, \mystate{2})$.}
  \label{fig:sim-max-illustration}
\end{figure}

Formally, a sim-max game consists of a set of states $\States$, a set of messages $\Messgs$ typically with much
fewer messages than states, a probability distribution $\Pr \in \Delta(\States)$ such that
$\Pr(\state)$ gives the probability that state $\state$ occurs, a similarity metric\footnote{A
  metric is a function of distance between any two points in a given space. It should satisfy
  certain axioms that ensure behavior that one would intuitively expect from the words
  ``similarity'' and ``distance'' alone, but these details do not matter for the purposes of
  this paper.}
  % : similarity should always be non-negative, zero only for the distance
  % between an element and itself, symmetric, and satisfy the triangle inequality that
  % $\similarity(x,z) \le \similarity(x,y) + \similarity(y,z)$
on states $\similarity \mycolon \States \times \States \rightarrow \mathds{R}$ such that
$\similarity(\mystate{1}, \mystate{2})$ is the (perceptual) similarity between $\mystate{1}$
and $\mystate{2}$, and a utility function $\Utils \mycolon \States \times \States \rightarrow
\mathds{R}$ such that $\Utils(\mystate{1}, \mystate{2})$ is the payoff for sender and receiver
for a play with actual state $\mystate{1}$ and receiver interpretation $\mystate{2}$. We
identify the receiver's acts with the states of the world, so that the game is one of guessing
the actual state, so to speak. The utility function should be a monotonically decreasing
function of similarity.
In its original formulation, a sender's behavior is represented by a pure strategy $\Spure \in \Messgs^\States$ which associates states to messages (given a state $\state \in \States$ which message $\messg \in \Messgs$ would be sent), and a receiver's behavior by a pure strategy $\Rpure \in \States^\Messgs$ which does the opposite.
A play in such a game is illustrated in Figure~\ref{fig:sim-max-illustration}.

@@

Fix a signaling game with finite states $\States$ and messages $\Messgs$. We consider an
interpretation game in which the receiver chooses states in $\States$ in response to
messages. Let $\Pr(\cdot) \in \Delta(\States)$ be the prior distribution over states and
$\Util \mycolon \States \times \Messgs \times \States \rightarrow \mathds{R}$ the utility
function shared by senders and receivers in the population. A \emph{behavioral strategy} is a
function that maps an agent's choice points to a probability distribution over available
choices.\footnote{Our focus is on behavioral strategies, not \emph{mixed strategies}, i.e.,
  probability distributions over functions from each choice point to an act, such as
  $s \in \Delta(\Messgs^{\States})$.  Dynamics on behavioral strategies assume that agents can
  adapt their behavior locally, i.e., independently at each choice point. Our
  focus on behavioral strategies greatly reduces the complexity of the dynamic and simplifies
  numerical simulations. But it also seems the more plausible choice for imitation-based update
  protocols: agents only observe how, on some occasion, some other agent behaved in \emph{one}
  particular situation, not how that agent would behave in \emph{all} relevant choice
  situations; they imitate the use of a single word, so to speak, not a whole lexicon. (See
  \citet{Cressman2003:Evolutionary-Dy} for more on the difference between dynamics on mixed or
  behavioral strategies.)} The sender's behavioral strategies are functions
$\Sstrat \in \Delta(\Messgs)^\States$, thus mapping each state $\state \in \States$ to a
probability of each message $\messg \in \Messgs$ being sent in $\state$; the receiver's are
functions $\Rstrat \in \Delta(\States)^\Messgs$, thus mapping each message $\messg \in \Messgs$
to a probability of each interpretation $\state \in \States$ being chosen in response to
$\messg$. Although behavioral strategies are probabilistic, evolutionary modeling usually
imagines that every individual agent has a non-probabilistic strategy. Behavioral strategies
then capture average population behavior. In a virtually infinite population, the number
$\Sstrat(\messg \probbar \state)$, for instance, is then the probability that a randomly
sampled sender would send message $\messg$ if the actual state was $\state$. Similarly,
$\Rstrat(\state \probbar \messg)$ is then the probability with which a randomly sampled
receiver interprets $\messg$ as $\state$.

The \emph{expected utility} of choices at each choice point is:\footnote{The notation
  $\propto$, for ``proportional to'', that is used here and hereafter means that the right-hand
  side must possibly yet be normalized so as to have probabilities sum to one. In general,
  writing $P(x) \propto f(x)$ for any function $f \mycolon X \rightarrow \mathds{R}$ is
  shorthand for $P(x) = \frac{f(x)}{\sum_{x' \in X} f(x')}$. }
\begin{align*}
  \EU(\messg, \state, \Rstrat) & = \sum_{\state \in \States}
  \Rstrat(\state' \probbar \messg) \myts U(\state, \messg, \state') \\
  \EU(\state', \messg, \Sstrat) & = \sum_{\state \in
    \States} P(\state \probbar \messg)  \myts
  U(\state, \messg, \state')\,, \text{\ \  where } &
P(\state \probbar \messg) & \propto \Pr(\state) \myts \Sstrat(\messg \probbar \state) \,.
\end{align*}
Expected utilities are sums, over all possible concrete outcomes, of the utilities of these
outcomes, weighted by how likely they are to occur. For instance, the expected utility for the
sender of choosing message $\messg$, given tate $\state$ and a receiver strategy $\Rstrat$ is
the sum, for each act $\act$, of the probability that some random receiver will choose
$\act$ given $\messg$ times the actual utility of triple $\state$, $\messg$ and $\act$.

The discrete-time replicator dynamic tracks changes in frequency of choices in the population
as proportional to their expected utilities
\citep[e.g.][]{HofbauerSigmund1998:Evolutionary-Ga}:\footnote{The formulation given in
  Equation~(\ref{eq:RDDiscrete}) is adequate only for cases like the one we will be looking at
  in Section~\ref{sec:exploring-rdd}, where utilities are always non-negative and expected
  utilities are always positive.}
\begin{align}
  \label{eq:RDDiscrete}
  \Sstrat'(\messg \probbar \state) & \propto \Sstrat(\messg \probbar \state) \cdot
    \EU(\messg, \state,\Rstrat)\,, & \Rstrat'(\act \probbar \messg) & \propto \Rstrat(\act \probbar \messg) \cdot
    \EU(\act, \messg,\Sstrat)\,.   
\end{align}
For a given choice point, say a state $\state$, the probability of seeing $\messg$ played by an
average agent in the population \emph{after} the update is proportional to the probability of
seeing it \emph{before} the update, which is $\Sstrat(\messg \probbar \state)$, times the
expected utility of $\messg$ at state $\state$. Intuitively put, frequencies of choices change
by a gradient of current frequency and a measure of how good they are.

\subsection{A noisy replicator dynamic}
\label{sec:noisy-repl-dynam}

The replicator dynamic is an abstract population-level dynamic that describes the mean expected
change of behavioral dispositions in a population of signalers. There are several ways of
deriving the replicator dynamic from agent-level processes of behavioral adaptation. We focus
here on one of the simplest: \emph{probabilistic imitation of success}
\citep[see][]{Sandholm2010:Population-Game}: agents imitate the behavior of some other agent,
which they happen to observe, with a probability that is proportional to the expected utility
of the latter agents' choice (see Appendix~XYZ for a derivation of the standard replicator
dynamic from this update scheme).

Probabilistic imitation of success presupposes that agents make no mistakes when observing
states. This may not always be an appropriate assumption. Especially for sim-max games, where,
intuitively, some states can be fairly similar and therefore likely to be confused for one
another. Appendix~XYZ shows how to derive a generalization of the replicator dynamic from the
assumption that probabilistic imitation of success is affected by two sources of probabilistic
confusion of states:

\begin{enumerate}
\item \emph{observation noise:} whenever a state $\mystate{a}$ actually occurs, the
  probability that is observed as $\mystate{o}$ is $P_o(\mystate{o} \probbar \mystate{a})$;
\item \emph{realization noise:} whenever a receiver intends to realize interpretation
  $\mystate{i}$, the probability that $\mystate{r}$ is realized is $P_r(\mystate{r} \probbar \mystate{o})$.
\end{enumerate}

The mean changes in the population frequencies of choices under noise-affected imitation of
success can be covered by the following discrete-time formulation:

\begin{align}
  \label{eq:nRD}
  \Sstrat'(\messg \probbar \state) & \propto P_o(\messg \probbar \state) \myts \EU(\messg,
  \state, \Rstrat)\,, & \Rstrat'(\state \probbar \messg) & \propto P_o(\state \probbar \messg)
  \myts \EU(\state, \messg, \Sstrat)\,.
\end{align}

This looks very much like the discrete-time formulation of the standard replicator dynamic in
(\ref{eq:RDDiscrete}), but there are differences (the Appendix~XYZ for full definitions and
formal detail). Firstly, expected utility here takes stochastic confusability of states into
account in a natural way. Secondly, where the standard replicator dynamic had probabilities
$\Sstrat(\messg \probbar \state)$ and $\Rstrat(\state \probbar \messg)$, we now have
$P_o(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ respectively. This is the major
difference, so we should dwell on it for a brief moment.

Look at the sender case first, where $\Sstrat(\messg \probbar \state)$ is the probability with
which $\messg$ is played in response to actual state $\state$ on average in the population. If
agents may not perceive the actual state, this should also affect what they observe when trying
to imitate another agents behavior. $P_o(\messg \probbar \state)$ is therefore the probability,
given that a potential imitator has observed $\state$, that he would observe a random agent
play $\messg$, which he may or may not choose to imitate. Similarly for the receiver, we are
not looking at occurrence probability $\Rstrat(\state \probbar \messg)$ but at the probability
$P_o(\state \probbar \messg)$ with which an agent who observes $\messg$ observes another agent
choose interpretation $\state$.

Obviously, $P_o(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ are to be derived
from, among others, probabilistic confusion of states $P_o(\mystate{o} \probbar \mystate{a})$
and $P_r(\mystate{r} \probbar \mystate{o})$. Details are given in Appendix~XYZ. If there is no
noise, $P_o(\messg \probbar \state)$ reduces to $\Sstrat(\messg \probbar \state)$ and
$P_o(\state \probbar \messg)$ reduces to $\Rstrat(\state \probbar \messg)$. But if there is,
imitation probabilities must take confusability of states into account. In this sense, the
noisy replicator dynamic in (\ref{eq:nRD}) conservatively extends the classic case in
(\ref{eq:RDDiscrete}).

@@

\appendix
